Modelling M/EEG data with Bayesian multilevel generalised additive models

### 2025-02-10 brms_meeg.tex
reshaping the document and drafting a script to conduct proper simulations of onset/offset recovery

\title{Modelling M/EEG data with Bayesian nonparametric multilevel
models}

@@

\abstract{Time-resolved electrophysoiological measurements such as those
offered by magneto- or electro-encephalography (M/EEG) provide a unique
window onto neural activity underlying cognitive process and how they
unfold over time. Typically, we are interested in testing whether such
measures differ across conditions and/or groups. The conventional
approach consists in conducting mass-univariate statistics followed by
some form of multiplicity correction (e.g., FDR, FWER) or cluster-based
inference. However, these cluster-based methods have an important
downside: they shift the focus of inference from the timepoint to the
cluster level, thus preventing any conclusion to be made about the onset
and offset of effects (e.g., differences across conditions). Here, we
introduce a novel \emph{model-based approch} for analysing
one-dimensional M/EEG timeseries such as ERPs or decoding timecourses
and their differences across conditions. This approach relies on
Bayesian nonparametric multilevel modelling (multilevel generalised
additive models or Gaussian processes), which outputs posterior
probabililities of the effect being above chance at every
timestep/voxel, while naturally taking into account the temporal
dependencies and between-subject variability present in such data.}

@@

\hypertarget{toc}{}
\tableofcontents
\newpage
\section[Introduction]{Modelling M/EEG data with Bayesian nonparametric
multilevel models}

@@

@@

Recent example of GLM for EEG (\citeproc{ref-fischer2013}{Fischer \&
Ullsperger, 2013}; \citeproc{ref-wuxfcllhorst2025}{Wüllhorst et al.,
2025})\ldots{} ``Similar approaches have been successfully applied to
EEG time- (Rousselet et al., 2008) and frequency-domain (Cohen and
Cavanagh, 2011) data and allow the simultaneous investigation of
multiple independent variables while preserving the high temporal
resolution of the EEG.''\ldots{} See also (\citeproc{ref-hauk2006}{Hauk
et al., 2006}; \citeproc{ref-rousselet2008}{Rousselet et al.,
2008})\ldots{} Example of two-stage regression analysis
(individual-level then group-level, 10.1101/2024.04.10.588861,
\citeproc{ref-dunagan2024}{Dunagan et al., 2024})\ldots{}

@@

\subsection{Gaussian process
regression}\label{gaussian-process-regression}

A Gaussian process (GP) is a stochastic process that defines the
distribution over a collection of random variables indexed by a
continuous variable, that is \(\{f(t): t \in \mathcal{T}\}\) for some
index set \(\mathcal{T}\) (\citeproc{ref-rasmussen2005}{Rasmussen \&
Williams, 2005};
\citeproc{ref-riutort-mayol_practical_2023}{Riutort-Mayol et al.,
2023}). Whereas Bayesian linear regression outputs a distribution over
the parameters of some predefind parametric model, the GP approach, in
contrast, is a non-parametric approach, in that it finds a distribution
over the possible functions that are consistent with the observed data.
However, note that nonparametric does not mean there aren't parameters,
it means that there are infinitely many parameters.

From
\href{https://www.rdocumentation.org/packages/brms/versions/2.22.0/topics/gp}{brms
documentation}: A GP is a stochastic process, which describes the
relation between one or more predictors
\(x=\left(x_{1}, \ldots, x_{d}\right)\) and a response \(f(x)\), where
\(d\) is the number of predictors. A GP is the generalization of the
multivariate normal distribution to an infinite number of dimensions.
Thus, it can be interpreted as a prior over functions. The values of
\(f()\) at any finite set of locations are jointly multivariate normal,
with a covariance matrix defined by the covariance kernel
\(k_p\left(x_i, x_j\right)\), where \(p\) is the vector of parameters of
the GP:

\[
\left(f\left(x_{1}\right), \ldots f\left(x_{n}\right) \sim \operatorname{MVN}\left(0, \left(k_p\left(x_{i}, x_{j}\right)\right)_{i, j=1}^{n}\right)\right.
\]

The smoothness and general behaviour of the function \(f\) depends only
on the choice of covariance kernel, which ensures that values that are
close together in the input space will be mapped to similar output
values\ldots{}

From this perspective, \(f\) is a realisation of an infinite dimensional
normal distribution:

\[
f \sim \mathrm{Normal}(0, C(\lambda))
\]

where \(C\) is a covariance kernel with hyperparameters \(\lambda\) that
defines the covariance between two function values \(f\left(t_1\right)\)
and \(f\left(t_2\right)\) for two time points \(t_1\) and \(t_2\)
(\citeproc{ref-rasmussen2005}{Rasmussen \& Williams, 2005}). Similar to
the different choices of the basis function for splines, different
choices of the covariance kernel lead to different GPs. In this article,
we consider the squared-exponential (a.k.a. radial basis function)
kernel, which computes the squared distance between points and converts
it into a measure of similarity. It is defined as:

\[
C(\lambda) := C\left(t_1, t_2, \sigma, \gamma\right) := \sigma^2 \exp \left(-\frac{||t_1-t_2||^{2}}{2 \gamma^2}\right)
\]

with hyperparameters \(\lambda = (\sigma, \gamma)\), expressing the
overall scale of GP and the length-scale, respectively
(\citeproc{ref-rasmussen2005}{Rasmussen \& Williams, 2005}). The
advantages of this kernel are that it is computationally efficient and
(infinitely) smooth making it a reasonable choice for the purposes of
the present article. Here again, \(\lambda\) hyperparameters are
estimated from the data, along with all other model parameters.

Taken from
\url{https://michael-franke.github.io/Bayesian-Regression/practice-sheets/10c-Gaussian-processes.html}:
For a given vector \(\mathbf{x}\), we can use the kernel to construct
finite multi-variate normal distribution associated with it like so:

\[
\mathbf{x} \mapsto_{G P} \operatorname{MVNormal}(m(\mathbf{x}), k(\mathbf{x}, \mathbf{x}))
\]

where \(m\) is a function that specifies the mean for the distribution
associated with \(\mathbf{x}\). This mapping is essentially the Gaussian
process: a systematic association of vectors of arbitrary length with a
suitable multi-variate normal distribution.

See also
\url{https://jtimonen.github.io/lgpr-usage/articles/math.html}\ldots{}

Low-rank approximate Gaussian processes are of main interest in machine
learning and statistics due to the high computational demands of exact
Gaussian process models
(\citeproc{ref-riutort-mayol_practical_2023}{Riutort-Mayol et al.,
2023})\ldots{}

@@

\caption{\label{fig-erp}Group-level average difference between
conditions (mean +/- standard error of the mean). The `true' (population
value of the) onset is indicated by the vertical dashed line.}

@@

We then fitted a GAM and GP regression model using the \texttt{brms}
package (\citeproc{ref-brms2017}{Bürkner, 2017},
\citeproc{ref-brms2018}{2018}; \citeproc{ref-nalborczyk2019}{Nalborczyk
et al., 2019})\ldots{}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

Now we fit the GP\ldots{}

@@

We can retrieve the posterior probability of the slope for
\texttt{condition} at each timestep.

\caption{Slope for the difference between conditions according to the
GAM (left) or the GP (right).}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/unnamed-chunk-1-1.pdf}

@@

We can also compute the posterior probability of the slope for
\texttt{condition} being above 0 (Figure~\ref{fig-post-prob-test}).

@@

\caption{\label{fig-post-prob-test}Posterior probability of the EEG
signal being above 0+eps according to the GAM (left) and the GP
(right).}

@@

We can also express this as the ratio of posterior probabilities (i.e.,
\(p/(1-p)\)) and visualise the timecourse of this ratio superimposed
with the conventional BF thresholds (Figure~\ref{fig-post-prob-ratio}).
Also check this paper
(\url{https://link.springer.com/article/10.3758/s13428-021-01613-6?fromPaywallRec=false})?

@@

\caption{\label{fig-post-prob-ratio}Ratio of posterior probability
according to the GAM (left) and the GP (right). Timesteps above
threshold (\textgreater10) are highlighted in green.}

@@

Below we are plotting the error function (where the error is computed as
\(\text{error}:=|\hat{\theta}-\theta|\)) for both the onset and offset
values of the ERP difference, according to various \texttt{eps} and
\texttt{threshold} values. Remember that the signal is generated from a
truncated Gaussian defining an objective onset at 160 ms, a maximum at
250 ms, and an offset at 342 ms. Figure~\ref{fig-onset-error} shows that
the GP model can \emph{exactly} recover the true onset and offset
values, given some reasonable choice of \texttt{eps} and
\texttt{threshold} values.

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.03         6            0.16            0.351           0        0.009
2 0.03         7            0.16            0.351           0        0.009
3 0.03         8            0.16            0.351           0        0.009
4 0.03         9            0.16            0.351           0        0.009
5 0.03        10            0.16            0.351           0        0.009
6 0.03        11            0.16            0.351           0        0.009
\end{verbatim}

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.07         8           0.166            0.342       0.006            0
2 0.07         9           0.166            0.342       0.006            0
3 0.07        10           0.166            0.342       0.006            0
4 0.07        11           0.166            0.342       0.006            0
5 0.07        12           0.166            0.342       0.006            0
6 0.07        13           0.166            0.342       0.006            0
\end{verbatim}

\begin{verbatim}
[1] eps              threshold        estimated_onset  estimated_peak  
[5] estimated_offset error_onset      error_peak       error_offset    
<0 rows> (or 0-length row.names)
\end{verbatim}

\caption{\label{fig-onset-error}Error function of onset (left) and
offset (right) estimation according to various eps and threshold values
(according to the GP model). Minimum error values are indicated by red
crosses.}

\subsection{Using summary statistics of
ERPs}\label{using-summary-statistics-of-erps}

Next we fit a hierarchical GAM using summary statistics of ERPs (mean
and SD) at the participant level (similar to what is done in
meta-analysis).

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\NormalTok{meta\_gam\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gam, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}
\NormalTok{meta\_gp\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gp, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}

@@

\subsection{Application to time-resolved decoding results (accuracy over
time)}\label{application-to-time-resolved-decoding-results-accuracy-over-time}

@@

@@

From Rousselet (2024): Five methods for multiple comparison correction
were considered: two FDR methods; two cluster-based methods; and the
maximum statistics. The two FDR methods were BH95 (Benjamini \&
Hochberg, 1995) and BY01 (Benjamini \& Yekutieli, 2001), which were
applied to the permutation p-values using the \texttt{p.adjust} function
in \texttt{R}. The first cluster-based inference was implemented using a
cluster-sum statistic of squared t-values\ldots{}

Comparing performance of the GAM and GP models with:

\begin{itemize}
\tightlist
\item
  uncorrected univariate tests
\item
  univariate tests + Bonferroni/Holm
\item
  univariate tests + BH95 (Benjamini-Hochberg, 1995) aka FDR
\item
  univariate tests + BY01 (Benjamini-Yekutieli, 2001)
\item
  cluster-based permutation test (cluster sum)
\item
  cluster-based permutation test (cluster depth)
\item
  threshold-free cluster enhancement
\item
  change point
\end{itemize}

We used the \texttt{R} package \texttt{changepoint} v 2.2.4
(\citeproc{ref-changepoint}{Killick et al., 2022a})\ldots{}

S-values (\citeproc{ref-greenland2019}{Greenland, 2019})\ldots{}

@@

@@

Extending the approach to spatiotemporal data (i.e., time +
sensors)\ldots{}

@@

@@

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting a GAM with two temporal dimensions}
\NormalTok{timegen\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# 2D thin{-}plate spline (tp)}
    \CommentTok{\# auc \textasciitilde{} s(train\_time, test\_time, bs = "tp", k = 10),}
\NormalTok{    auc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{t2}\NormalTok{(train\_time, test\_time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ timegen\_data,}
    \AttributeTok{family =} \FunctionTok{Beta}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/timegen\_gam\_t2.rds"}
\NormalTok{    )}

>>>>

\title{Modelling M/EEG data with Bayesian multilevel generalised
additive models}

@@

\abstract{Time-resolved electrophysiological measurements such as those
offered by magneto- or electro-encephalography (M/EEG) provide a unique
window onto neural activity underlying cognitive process and how they
unfold over time. Typically, we are interested in testing whether such
measures differ across conditions and/or groups. The conventional
approach consists in conducting mass-univariate statistics through
followed by some form of multiplicity correction (e.g., FDR, FWER) or
cluster-based inference. However, these cluster-based methods have an
important downside: they shift the focus of inference from the timepoint
to the cluster level, thus preventing any conclusion to be made about
the onset and offset of effects (e.g., differences across conditions).
Here, we introduce a novel \emph{model-based approch} for analysing
one-dimensional M/EEG timeseries such as ERPs or decoding timecourses
and their differences across conditions or group. This approach relies
on Bayesian nonparametric multilevel modelling (multilevel generalised
additive models or Gaussian processes), which outputs the posterior
probabilility of the effect being different than 0 (or above chance) at
every timestep, while naturally taking into account the temporal
dependencies and between-subject variability present in such data.}

@@

\hypertarget{toc}{}
\tableofcontents
\newpage
\section[Introduction]{Modelling M/EEG data with Bayesian multilevel
generalised additive models}

@@

From Dimigen \& Ehinger (\citeproc{ref-dimigen2021}{2021}): ``Recently,
spline regression has been applied to ERPs (Hendrix, Baayen, \& Bolger,
2017; Kryuchkova,Tucker,Wurm,\&Baayen,2012;Tremblay\& Baayen, 2010;
Tremblay \& Newman, 2015)''\ldots{}

@@

Recent example of GLM for EEG (\citeproc{ref-fischer2013}{Fischer \&
Ullsperger, 2013}; \citeproc{ref-wuxfcllhorst2025}{Wüllhorst et al.,
2025})\ldots{} ``Similar approaches have been successfully applied to
EEG time- (Rousselet et al., 2008) and frequency-domain (Cohen and
Cavanagh, 2011) data and allow the simultaneous investigation of
multiple independent variables while preserving the high temporal
resolution of the EEG.''\ldots{} See also (\citeproc{ref-hauk2006}{Hauk
et al., 2006}; \citeproc{ref-rousselet2008}{Rousselet et al.,
2008})\ldots{} Example of two-stage regression analysis (i.e.,
individual-level then group-level, \citeproc{ref-dunagan2024}{Dunagan et
al., 2024})\ldots{}

@@

@@

\caption{\label{fig-erp}Group-level average difference between
conditions (mean +/- standard error of the mean). The `true' onset and
offset are indicated by the vertical dashed lines.}

@@

We then fitted a GAM and GP regression model using the \texttt{brms}
package (\citeproc{ref-brms2017}{Bürkner, 2017},
\citeproc{ref-brms2018}{2018}; \citeproc{ref-nalborczyk2019}{Nalborczyk
et al., 2019}). We used the default priors in \texttt{brms}, that is,
weakly informative priors. Four Markov Chain Monte-Carlo (MCMC) were ran
for each model to approximate the posterior distribution, including each
5000 iterations and a warmup of 2000 iterations. Posterior convergence
was assessed examining trace plots as well as the Gelman--Rubin statisti
\(\hat{R}\).

@@

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

Now we fit the GP regression model\ldots{}

@@

We can retrieve the posterior probability of the slope for
\texttt{condition} at each timestep (Figure~\ref{fig-plot-post-slope}).

\caption{\label{fig-plot-post-slope}Slope for the difference between
conditions according to the GAM (left) or the GP (right).}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-plot-post-slope-1.pdf}

@@

We can also compute the posterior probability of the slope for
\texttt{condition} being above \(0+\epsilon\)
(Figure~\ref{fig-post-prob-test}), with \(\epsilon\) defined as 10\% of
the standard deviation of the raw EEG signal.

@@

\caption{\label{fig-post-prob-test}Posterior probability of the ERP
difference (slope) being above 0 according to the GAM (left) and the GP
(right).}

@@

We can also express this as the ratio of posterior probabilities (i.e.,
\(p/(1-p)\)) and visualise the timecourse of this ratio superimposed
with the conventional thresholds on evidence ratios
(Figure~\ref{fig-post-prob-ratio}).

@@

\caption{\label{fig-post-prob-ratio}Ratio of posterior probability
according to the GAM. Timesteps above threshold (10) are highlighted in
green.}

@@

\subsection{Multilevel modelling using ERP summary
statistics}\label{multilevel-modelling-using-erp-summary-statistics}

Next we fit a hierarchical/multilevel GAM using summary statistics of
ERPs (mean and SD) at the participant level (similar to what is done in
meta-analysis).

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\NormalTok{meta\_gam\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gam, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{scale\_colour\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{scale\_fill\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}
\NormalTok{meta\_gp\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gp, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{scale\_colour\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{scale\_fill\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}

@@

We then computed the difference between the true and estimated
onset/offset of the ERP difference
(\(\text{error}:=|\hat{\theta}-\theta|\)), according to various
\texttt{eps} and \texttt{threshold} values. Remember that the signal is
generated from a truncated Gaussian with an objective onset at 160 ms, a
maximum at 250 ms, and an offset at 342 ms. Figure~\ref{fig-onset-error}
shows that the hierarchical GAM can \emph{exactly} recover the true
onset and offset values, given some reasonable choice of \texttt{eps}
and \texttt{threshold} values (e.g., a threshold of 20).

\begin{verbatim}
   eps threshold estimated_onset estimated_offset error_onset error_offset
1 0.00        13            0.16             0.34           0        0.002
2 0.00        14            0.16             0.34           0        0.002
3 0.00        15            0.16             0.34           0        0.002
4 0.00        16            0.16             0.34           0        0.002
5 0.00        17            0.16             0.34           0        0.002
6 0.01        10            0.16             0.34           0        0.002
\end{verbatim}

@@

\caption{\label{fig-onset-error}Error function of onset (left) and
offset (right) estimation according to various eps and threshold values
(according to the hierarchical GAM). Minimum error values are indicated
by red crosses.}

@@

We considered five methods for multiplicity corrections: two FDR
methods; two cluster-based methods; and the maximum statistics. The two
FDR methods were BH95 (Benjamini \& Hochberg, 1995;
10.1111/j.2517-6161.1995.tb02031.x) and BY01 (Benjamini \& Yekutieli,
2001; 10.1214/aos/1013699998), which were applied to the permutation
p-values using the \texttt{p.adjust} function in \texttt{R}. The first
cluster-based inference was implemented using a cluster-sum statistic of
squared t-values\ldots{}

We also compared the performance of the GAMM (in correctly
identifying/estimating the onset and offset of ERP difference across
conditions) to the performance of the binary segmentation method as
implemented in the \texttt{R} package \texttt{changepoint} v 2.2.4
(\citeproc{ref-changepoint}{Killick et al., 2022a}), threshold-free
cluster enhancement (TFCE, \citeproc{ref-smith2009}{Smith \& Nichols,
2009}), as implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), and using the raw t-, p-,
or s-values timecourse. S-values
(\citeproc{ref-greenland2019}{Greenland, 2019})\ldots{}

@@

\subsection{Simulation study}\label{simulation-study}

Onset/offset estimation methods were assessed using these summary
statistics of the Monte Carlo sampling distributions of onsets: bias
(median-based), mean absolute error (MAE) and variance, with 10.000
iterations (i.e., 10.000 simulated datasets)\ldots{}

@@

Can be applied to any 1D timeseries (e.g., pupillometry,
electromyography)\ldots{} Extending the approach to spatiotemporal data
(i.e., time + sensors)\ldots{}

@@

\bibitem[\citeproctext]{ref-dimigen2021}
Dimigen, O., \& Ehinger, B. V. (2021). Regression-based analysis of
combined EEG and eye-tracking data: Theory and applications.
\emph{Journal of Vision}, \emph{21}(1), 3.
\url{https://doi.org/10.1167/jov.21.1.3}

@@

\section{Application to time-resolved decoding results (accuracy over
time)}\label{application-to-time-resolved-decoding-results-accuracy-over-time}

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting a GAM with two temporal dimensions}
\NormalTok{timegen\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# 2D thin{-}plate spline (tp)}
    \CommentTok{\# auc \textasciitilde{} s(train\_time, test\_time, bs = "tp", k = 10),}
\NormalTok{    auc }\SpecialCharTok{\textasciitilde{}} \FunctionTok{t2}\NormalTok{(train\_time, test\_time, }\AttributeTok{bs =} \StringTok{"tp"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ timegen\_data,}
    \AttributeTok{family =} \FunctionTok{Beta}\NormalTok{(),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/timegen\_gam\_t2.rds"}
\NormalTok{    )}
